Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-11 concurrency/CONCUR 2023/paper.tex
fix breakage

% comments from authors 
\newcommand{\than}[1]{\textbf{\textcolor{blue}{[ Than: #1]}}}
\newcommand{\lb}[1]{\textbf{\textcolor{red}{[ Lennart: #1]}}}

\newcommand{\ignore}[1]{}

@@

@@

\section{Search Structure Templates}
The C implementation of a node in our template style is:
\begin{lstlisting}[language = C, backgroundcolor=\color{white},   
	commentstyle=\color{codegreen},
	basicstyle=\ttfamily\footnotesize,
	breakatwhitespace=false,  breaklines=true,  keepspaces=true,                                
	showspaces=false,  showstringspaces=false,
	showtabs=false, tabsize=1]
typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
typedef struct pn {struct tree_t *p; struct tree_t *n;} pn;
\end{lstlisting}
Each node of type \lstinline{tree_t} has a \lstinline{lock} field that protects the node and a \lstinline{t} field that points to a \lstinline{tree} struct. The \lstinline{t} field can be either \lstinline{NULL}, indicating that the node is a leaf node with no key or value, or it can point to a \lstinline{tree} struct that stores the node's key, value, and child pointers.

The \lstinline{pn} struct contains two nodes of type \lstinline{tree_t}: the \lstinline{p} field, which denotes the current node, and the \lstinline{n} field, which represents the next node. To traverse the tree using hand-over-hand locking, we must maintain at least one lock during traversal between nodes. This means that the lock for the next node is obtained before releasing the lock for the current node.

@@

To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}. The \lstinline{pn} struct includes two fields, namely \lstinline{p} and \lstinline{n}, which respectively represent the current and next node. Additionally, we require a target key \lstinline{k} and a dummy node value \lstinline{value} (referred to as 'dummy' because it is not actually used in \lstinline{traverse}). We include this dummy node to maintain consistency with other implementations. Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findnext}. The purpose of \lstinline{findnext} is to determine the next node to be visited during the traversal. In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}. Alternatively, it terminates if \lstinline{findnext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values. The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal. If it fails to do so, it returns false, which other functions can use to make decisions.

@@

When \lstinline{findnext} arrives at a new node, it acquires a lock for that node, as is the nature of hand-over-hand locking. It then releases the lock for the current node, ensuring that the link between the two nodes is not removed or rearranged while traversing it. To utilize the template methodology for concurrent data structure implementations, it is crucial to keep track of the locks acquired for each node during the traversal.

@@

%% C implementation 
To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node. Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{tree_t} in our C implementation. These fields store the lower and upper bounds of the keys, respectively. When a thread arrives at a node in search of the key $\texttt{k}$, it checks whether $\texttt{k}$ falls within the range of that node. If $\texttt{k}$ falls within the range, the thread continues by calling \lstinline{findnext} to find the new node. If $\texttt{k}$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).

\subsection{Templates vs. Internal Reorganization}

@@

$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} 
%\ast \lockinv \texttt{(pn->p->lock,R'}) 
\ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
$$\texttt{traverse(pn, k)}$$
$$\left\langle \texttt{n'}.\ \nodeboxrep \texttt{(pn)} \
%ast \lockinv \texttt{(pn->n'->lock,R')} 
\ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range})
\ |\ \treerep\ \texttt{t} \right\rangle$$

@@

>>>>

% comments from authors 
\newcommand{\than}[1]{\textbf{\textcolor{blue}{[Than: #1]}}}
\newcommand{\lb}[1]{\textbf{\textcolor{red}{[Lennart: #1]}}}
\newcommand{\wm}[1]{\textbf{\textcolor{violet}{[Willam: #1]}}}
\newcommand{\ignore}[1]{}

@@

\wm{point 1 - explain the top-level goal and approach of templates, how it relates to modularity}

\wm{point 2 - lay out the template approach, what a template is, what a data structure is, how their code and their specifications interface}

@@

\section{Search Structure Templates}
The C implementation of a node in our template style is:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node_t {node *t; lock_t *lock;} node_t;
typedef struct pn {struct node_t *p; struct node_t *n;} pn;
\end{lstlisting}
\than{I changed \lstinline{tree_t} into \lstinline{node_t}, and \lstinline{tree} into \lstinline{node}, since concurrent search structure will not depend on particular data structures.}

Each node of type \lstinline{node_t} has a \lstinline{lock} field that protects the node and a \lstinline{t} field that points to a \lstinline{node} struct. Depending on the data structure, the \lstinline{t} field can have different interpretations. In a binary search tree, if the \lstinline{t} field is \lstinline{NULL}, it indicates a leaf node with no key or value; otherwise, it points to a \lstinline{node} struct containing the node's key, value, and child pointers. 
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node {int key; void *value; struct node_t *left, *right;} node;
\end{lstlisting}

For a linked list, a \lstinline{t} field with \lstinline{NULL} represents the last node in the list, while a non-null \lstinline{t} field points to a \lstinline{node} struct holding the node's key, value, and a pointer to the next item in the list.

\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node {int key; void *value; struct node_t *next;} node;
\end{lstlisting}

For the purpose of reasoning about lock-coupling style (also known as hand-over-hand locking), we specifically consider traversing the tree using this locking method. To accomplish this, we must maintain at least one lock during traversal between nodes. This requires obtaining the lock for the next node before releasing the lock for the current node. Therefore, we utilize the \lstinline{pn} struct containing two \lstinline{node_t} type nodes: the \lstinline{p} field representing the current node, and the \lstinline{n} field denoting the next node.

@@

To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}. The \lstinline{pn} struct includes two fields, namely \lstinline{p} and \lstinline{n}, which respectively represent the current and next node. Additionally, we require a target key \lstinline{k} and a dummy node value \lstinline{value} (referred to as 'dummy' because it is not actually used in \lstinline{traverse}). We include this dummy node to maintain consistency with other implementations. Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findNext}. The purpose of \lstinline{findNext} is to determine the next node to be visited during the traversal. In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}. Alternatively, it terminates if \lstinline{findNext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values. The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal. If it fails to do so, it returns false, which other functions can use to make decisions.

@@

When \lstinline{findNext} arrives at a new node, it acquires a lock for that node, as is the nature of hand-over-hand locking. It then releases the lock for the current node, ensuring that the link between the two nodes is not removed or rearranged while traversing it. To utilize the template methodology for concurrent data structure implementations, it is crucial to keep track of the locks acquired for each node during the traversal.

@@

%% C implementation 
To implement the template algorithm, it is crucial to keep track of the lower and upper bounds of the keys present in the subtree rooted at the current node. Therefore, we have incorporated two new fields, \lstinline{min} and \lstinline{max}, into each node of type \lstinline{tree_t} in our C implementation. These fields store the lower and upper bounds of the keys, respectively. When a thread arrives at a node in search of the key $\texttt{k}$, it checks whether $\texttt{k}$ falls within the range of that node. If $\texttt{k}$ falls within the range, the thread continues by calling \lstinline{findNext} to find the new node. If $\texttt{k}$ does not fall within the range, the thread releases the lock and returns to the root node by assigning the pointer of the root to the pointer of the current node (lines 21-22 in Figure \ref{traverse_giveup_a}).

\subsection{Templates vs. Internal Reorganization} % move to section 4?
In the binary search tree, \texttt{pushdown\_left} involves both the structure of the tree and the concurrency protocol. There's no good way to disentangle this into a template and a concurrency-free data structure operation.

@@

$$\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn)} 
%\ast \lockinv \texttt{(pn->p->lock,R'}) 
\ast \texttt{R(pn->p)} \ | \ \treerep\ \texttt{t} \right\rangle $$
$$\texttt{traverse(pn, k)}$$
$$\left\langle \texttt{res}, \texttt{n'}. \ \nodeboxrep \texttt{(pn)}
%ast \lockinv \texttt{(pn->n'->lock,R')} 
\ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range}) \ast (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))
\   |\ \treerep\ \texttt{t} \right\rangle$$

@@

\section{Proof Mechanization}
